---
title: "Filter variants by predicted effect, wecare feb2016"
author: "Alexey Larionov"
output: html_document
params:
  interim_data: ""
  priority_genes: ""
---

started: 01Mar2016  
last updated: 09Aug2016

# Criteria

### Loss of function: 
all (~5k)

### ClinSig
non-benign (~1k)

### COSM in Existing_Variation
all (~28k)

### Priority genes (in-house lists):
BC risk (65), BC somatic(94), DNA repair (143) and Estrogen-signalling related (7)

### Missense

*strict*: deleterious and probably_damaging with max scores (~2k)  
*medium_1*: deleterious and probably_damaging with any scores (~14k)  
*medium_2*: deleterious or probably_damaging with any scores (~34k)  
*relaxed*: (all missence): ~81k  

*priority_strict*: strict + priority_gene (146)  
*priority_relaxed*: relaxed + priority_gene (1540)  

# start_section

```{r start_section}

# Start time
Sys.time()

# Do NOT clean-up at this stage
#rm(list=ls()) # this would remove params object!
graphics.off()

# Data folder
interim_data_folder <- params$interim_data
priority_genes_folder <- params$priority_genes

library(dplyr)

# For debugging
#interim_data_folder <- "/scratch/medgen/users/alexey/wecare_aug2016/interim_data"
#priority_genes_folder <- "/scratch/medgen/users/alexey/wecare_aug2016/gene_lists"
#setwd("/scratch/medgen/scripts/wecare_skat_08.16/scripts")

```

# load_and_check_data

```{r load_and_check_data}

# Remove params object to avoid interference with load() function
rm(params)

interim_data_file <- paste(interim_data_folder, "s03_filter_genotypes_feb2016.RData", sep="/")
load(file=interim_data_file)
ls()

dim(covar.df)
str(covar.df)
covar.df[1:5,1:5]

dim(demographics.df)
str(demographics.df)
demographics.df[1:5,1:5]

dim(samples.df)
str(samples.df)
samples.df[1:5,]

dim(vv.df)
str(vv.df)
vv.df[1:5,1:5]

dim(gt.mx)
class(gt.mx)
gt.mx[1:15,1:5]

dim(gq.mx)
class(gq.mx)
gq.mx[1:15,1:5]

dim(dp.mx)
class(dp.mx)
dp.mx[1:15,1:5]

dim(ref.mx)
class(ref.mx)
ref.mx[1:15,1:5]

dim(alt.mx)
class(alt.mx)
alt.mx[1:15,1:5]

# Check consistence of rownames
sum(rownames(gt.mx) != rownames(vv.df), na.rm=TRUE)

sum(rownames(gt.mx) != rownames(gq.mx), na.rm=TRUE)
sum(rownames(gt.mx) != rownames(dp.mx), na.rm=TRUE)
sum(rownames(gt.mx) != rownames(ref.mx), na.rm=TRUE)
sum(rownames(gt.mx) != rownames(alt.mx), na.rm=TRUE)

# Check consistence of colnames (sub(".DP","") may be used instead)
sum(substr(colnames(gt.mx),0,6) != substr(colnames(gq.mx),0,6), na.rm=TRUE)
sum(substr(colnames(gt.mx),0,6) != substr(colnames(dp.mx),0,6), na.rm=TRUE)
sum(substr(colnames(gt.mx),0,6) != substr(colnames(ref.mx),0,6), na.rm=TRUE)
sum(substr(colnames(gt.mx),0,6) != substr(colnames(alt.mx),0,6), na.rm=TRUE)

```

# explore_annotations

SIFT: tolerated, deleterious, tolerated_low_confidence, deleterious_low_confidence

PolyPhen: benign, possibly_damaging, probably_damaging, unknown          

CLIN_SIG: likely_benign, benign, risk_factor, not_provided, uncertain_significance, 
pathogenic, likely_pathogenic, association, protective, other, drug_response,  confers_sensitivity

Somatic: COSM in Existing_variation

```{r explore_annotations}

# SIFT
unique(sub("\\(.*\\)","",as.vector(vv.df$SIFT)))

# PolyPhen
unique(sub("\\(.*\\)","",as.vector(vv.df$PolyPhen)))

#CLIN_SIG
unique(unlist(strsplit(as.vector(vv.df$CLIN_SIG), split="\\&")))

```

# split_calls_and_scores_in_SIFT

SIFT: the higher score the less deleterious

```{r split_calls_and_scores_in_SIFT}

SIFT_Call <- 
  vv.df$SIFT  %>% 
  as.vector() %>% 
  sub("\\(.*\\)","",.)

NA -> SIFT_Call[SIFT_Call == ""]

length(SIFT_Call)
SIFT_Call[c(1:10,49)]

SIFT_Score <- 
  vv.df$SIFT %>% 
  as.vector() %>% 
  sub(".*\\(","",.)  %>% 
  sub("\\)","",.)

NA -> SIFT_Score[SIFT_Score == ""]
SIFT_Score <- as.numeric(SIFT_Score)

length(SIFT_Score)
SIFT_Score[1:10]

vv.df <- cbind(vv.df, SIFT_Call, SIFT_Score)
rm (SIFT_Call, SIFT_Score)
vv.df[c(1:10,180:190),c("SIFT", "SIFT_Call", "SIFT_Score")]

```

# split_calls_and_scores_in_PolyPhen

PolyPhen: the higher score the more damaging

```{r split_calls_and_scores_in_PolyPhen}

PolyPhen_Call <- 
  vv.df$PolyPhen  %>% 
  as.vector() %>% 
  sub("\\(.*\\)","",.)

NA -> PolyPhen_Call[PolyPhen_Call == ""]

length(PolyPhen_Call)
PolyPhen_Call[1:10]

PolyPhen_Score <- 
  vv.df$PolyPhen %>% 
  as.vector() %>% 
  sub(".*\\(","",.)  %>% 
  sub("\\)","",.)

NA -> PolyPhen_Score[PolyPhen_Score == ""]
PolyPhen_Score <- as.numeric(PolyPhen_Score)

length(PolyPhen_Score)
PolyPhen_Score[1:10]

vv.df <- cbind(vv.df, PolyPhen_Call, PolyPhen_Score)
rm (PolyPhen_Call, PolyPhen_Score)
vv.df[c(1:10,49),c("PolyPhen", "PolyPhen_Call", "PolyPhen_Score")]

```

# make_lof_index

```{r make_lof_index}

splice_acceptor_variant <- grepl( "splice_acceptor_variant", vv.df$Consequence)
sum(splice_acceptor_variant)

splice_donor_variant <- grepl( "splice_donor_variant", vv.df$Consequence)
sum(splice_donor_variant)

stop_gain <- grepl( "stop_gain", vv.df$Consequence)
sum(stop_gain)

frameshift_variant <- grepl( "frameshift_variant", vv.df$Consequence)
sum(frameshift_variant)

stop_lost <- grepl( "stop_lost", vv.df$Consequence)
sum(stop_lost)

start_lost <- grepl( "start_lost", vv.df$Consequence)
sum(start_lost)

lof_all <- splice_acceptor_variant | splice_donor_variant | stop_gain | frameshift_variant | stop_lost | start_lost

sum(lof_all)

rm(splice_acceptor_variant, splice_donor_variant, stop_gain, frameshift_variant, stop_lost, start_lost)

```

# make_missense_indices

```{r make_missense_indices}

missense_all <- grepl( "missense_variant", vv.df$Consequence)
sum(missense_all)

deleterious <- vv.df$SIFT_Call == "deleterious"
FALSE -> deleterious[is.na(deleterious)]
sum(deleterious)

probably_damaging <- vv.df$PolyPhen_Call == "probably_damaging"
FALSE -> probably_damaging[is.na(probably_damaging)]
sum(probably_damaging)

missense_medium_1 <- deleterious & probably_damaging
missense_medium_2 <- deleterious | probably_damaging
sum(missense_medium_1)
sum(missense_medium_2)
rm(deleterious, probably_damaging)

deleterious_strict <- vv.df$SIFT_Call == "deleterious" & vv.df$SIFT_Score == 0
FALSE -> deleterious_strict[is.na(deleterious_strict)]
sum(deleterious_strict)

damaging_strict <- vv.df$PolyPhen_Call == "probably_damaging" & vv.df$PolyPhen_Score == 1
FALSE -> damaging_strict[is.na(damaging_strict)]
sum(damaging_strict)

missense_strict <- deleterious_strict & damaging_strict
sum(missense_strict)
rm(deleterious_strict, damaging_strict)

```

# make_clinsig_index

Note sums w/o na.rm=TRUE

```{r make_clinsig_index}

benign <- grepl( "benign", vv.df$CLIN_SIG) # includes "likely_benign"
sum(benign)

risk_factor <- grepl( "risk_factor", vv.df$CLIN_SIG)
sum(risk_factor)

pathogenic <- grepl( "pathogenic", vv.df$CLIN_SIG) # includes "likely_pathogenic"
sum(pathogenic)

association <- grepl( "association", vv.df$CLIN_SIG)
sum(association)

protective <- grepl( "protective", vv.df$CLIN_SIG)
sum(protective)

drug_response <- grepl( "drug_response", vv.df$CLIN_SIG)
sum(drug_response)

confers_sensitivity <- grepl( "confers_sensitivity", vv.df$CLIN_SIG)
sum(confers_sensitivity)

other <- grepl( "other", vv.df$CLIN_SIG)
sum(other)

clinsig_non_benign <- risk_factor | pathogenic | association | protective | drug_response | confers_sensitivity | other

sum(clinsig_non_benign)

rm(benign, risk_factor, pathogenic, association, protective, drug_response, confers_sensitivity, other)

```

# make_somatic_index

```{r make_somatic_index}

somatic <- grepl("COSM", vv.df$Existing_variation)
sum(somatic)

```

# make_priority_genes_index

```{r make_priority_genes_index}

# Read gene lists
bc_risk_file <- paste(priority_genes_folder, "bc_risk.txt", sep="/")
bc_risk_genes <- read.table(bc_risk_file)
bc_risk_genes <- as.vector(bc_risk_genes[,1])
rm(bc_risk_file)

bc_somatic_file <- paste(priority_genes_folder, "bc_somatic.txt", sep="/")
bc_somatic_genes <- read.table(bc_somatic_file)
bc_somatic_genes <- as.vector(bc_somatic_genes[,1])
rm(bc_somatic_file)

dna_repair_file <- paste(priority_genes_folder, "dna_repair.txt", sep="/")
dna_repair_genes <- read.table(dna_repair_file)
dna_repair_genes <- as.vector(dna_repair_genes[,1])
rm(dna_repair_file)

es_related_file <- paste(priority_genes_folder, "es_related.txt", sep="/")
es_related_genes <- read.table(es_related_file)
es_related_genes <- as.vector(es_related_genes[,1])
rm(es_related_file)

# Make separate indices
bc_risk <- vv.df$SYMBOL %in% bc_risk_genes
bc_somatic <- vv.df$SYMBOL %in% bc_somatic_genes
dna_repair <- vv.df$SYMBOL %in% dna_repair_genes
es_related <- vv.df$SYMBOL %in% es_related_genes

# Add indices to vv.df
vv.df <- cbind(vv.df, bc_risk, bc_somatic, dna_repair, es_related)
x <- c("TP53", "ERCC1")
x.rows <- vv.df$SYMBOL %in% x
vv.df[x.rows,c("SYMBOL", "bc_risk", "bc_somatic", "dna_repair", "es_related")]
rm(x, x.rows)

# Make cumulative index
priority_genes <- bc_risk | bc_somatic | dna_repair | es_related
sum(priority_genes) # 4,063

# Clean-up
rm(bc_risk, bc_somatic, dna_repair, es_related, bc_risk_genes, bc_somatic_genes, dna_repair_genes, es_related_genes, priority_genes_folder)
  
```

# make_cumulative_indices

```{r make_cumulative_indices}

strict_set <- lof_all | missense_strict | clinsig_non_benign
sum(strict_set) # ~7k

standard_set_1 <- lof_all | missense_medium_1 | clinsig_non_benign
sum(standard_set_1) # ~18k

standard_set_2 <- lof_all | missense_medium_2 | clinsig_non_benign
sum(standard_set_2) # ~38k

relaxed_set <- lof_all | missense_all | clinsig_non_benign | somatic
sum(relaxed_set) # ~95k

priority_strict <- strict_set & priority_genes
sum(priority_strict) # 146

priority_relaxed <- relaxed_set & priority_genes
sum(priority_relaxed) # 1540

rm(lof_all, missense_strict, missense_medium_1, missense_medium_2, missense_all, clinsig_non_benign, somatic, priority_genes)

```

# filter_variants

```{r filter_variants}

vv.str <- vv.df[strict_set,]
gt.str <- gt.mx[strict_set,]
gq.str <- gq.mx[strict_set,]
dp.str <- dp.mx[strict_set,]
alt.str <- alt.mx[strict_set,]
ref.str <- ref.mx[strict_set,]

dim(gt.str)
dim(vv.str)
rm(strict_set)

vv.std.1 <- vv.df[standard_set_1,]
gt.std.1 <- gt.mx[standard_set_1,]
gq.std.1 <- gq.mx[standard_set_1,]
dp.std.1 <- dp.mx[standard_set_1,]
alt.std.1 <- alt.mx[standard_set_1,]
ref.std.1 <- ref.mx[standard_set_1,]

dim(gt.std.1)
dim(vv.std.1)
rm(standard_set_1)

vv.std.2 <- vv.df[standard_set_2,]
gt.std.2 <- gt.mx[standard_set_2,]
gq.std.2 <- gq.mx[standard_set_2,]
dp.std.2 <- dp.mx[standard_set_2,]
alt.std.2 <- alt.mx[standard_set_2,]
ref.std.2 <- ref.mx[standard_set_2,]

dim(gt.std.2)
dim(vv.std.2)
rm(standard_set_2)

vv.rel <- vv.df[relaxed_set,]
gt.rel <- gt.mx[relaxed_set,]
gq.rel <- gq.mx[relaxed_set,]
dp.rel <- dp.mx[relaxed_set,]
alt.rel <- alt.mx[relaxed_set,]
ref.rel <- ref.mx[relaxed_set,]

dim(gt.rel)
dim(vv.rel)
rm(relaxed_set)

vv.pri.rel <- vv.df[priority_relaxed,]
gt.pri.rel <- gt.mx[priority_relaxed,]
gq.pri.rel <- gq.mx[priority_relaxed,]
dp.pri.rel <- dp.mx[priority_relaxed,]
alt.pri.rel <- alt.mx[priority_relaxed,]
ref.pri.rel <- ref.mx[priority_relaxed,]

dim(gt.pri.rel)
dim(vv.pri.rel)
rm(priority_relaxed)

vv.pri.str <- vv.df[priority_strict,]
gt.pri.str <- gt.mx[priority_strict,]
gq.pri.str <- gq.mx[priority_strict,]
dp.pri.str <- dp.mx[priority_strict,]
alt.pri.str <- alt.mx[priority_strict,]
ref.pri.str <- ref.mx[priority_strict,]

dim(gt.pri.str)
dim(vv.pri.str)
rm(priority_strict)

```

# explore_result

POLG & POLQ are DNA-repair genes, previously not implicated in breast cancer  
ECT2L is a gene somatically mutated in BC, previously not associated with BC risk  
WRN is a DNA-repair gene, previously not implicated in breast cancer  

```{r explore_result}

# Explore selected genes 
# Note that sum of AFs per gene may exceed 1

vv.std.1[vv.std.1$SYMBOL=="NPHP4",c("SYMBOL","AF","AC","AN")]
vv.std.1[vv.std.1$SYMBOL=="OR4C5",c("SYMBOL","AF","AC","AN")]
vv.std.1[vv.std.1$SYMBOL=="SARM1",c("SYMBOL","AF","AC","AN")]

vv.pri.str[vv.pri.str$SYMBOL=="ECT2L",c("SYMBOL","AF","AC","AN")]
vv.pri.str[vv.pri.str$SYMBOL=="POLG",c("SYMBOL","AF","AC","AN")]
sort(table(as.vector(vv.pri.str$SYMBOL)), decreasing = TRUE)

#View(vv.pri.str[vv.pri.str$SYMBOL=="POLQ",])

```

# save_data

```{r save_data}

interim_data_file <- paste(interim_data_folder, "s04_filter_by_effect_feb2016.RData", sep="/")
save.image(file=interim_data_file)

```

# final_section

```{r final_section}

sessionInfo()
Sys.time()

```
